Viral and Bacterial Communities of Colorectal Cancer

### 2017-03-28 doc/manuscript.Rmd
Huge updates incorporating Pats comments.

Colorectal cancer is the second leading cause of cancer-related death in the United States and is a primary cause of morbidity and mortality throughout the world. Although etiology for the majority of colorectal cancer cases remain unclear, its progression has been linked to changes in colonic bacterial community composition. Viruses are another important component of the colonic microbial community. However, they have yet to be studied in colorectal cancer, despite their oncogenic potential. We evaluated the colorectal cancer virome (virus community) using a cohort of 90 human subjects with either healthy, adenomatous (precancerous), or cancerous colons. We utilized 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing methods to compare the colorectal cancer virome to the bacterial community. We found that alpha and beta diversity metrics were insufficient for detecting virome changes in colorectal cancer, but more sophisticated random forest models identified striking changes in the virus community. The majority of the cancer-associated virome consisted of temperate bacteriophages, suggesting that the community was indirectly linked to colorectal cancer by modulating bacterial community structure and function. These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and likely impact cancer progression by altering the bacterial host communities.

**Word Count**: 201 / 250

@@

# Introduction
Due to their mutagenic abilities and propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk], and because bacteria have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb], bacteriophages have the potential to indirectly impact cancer. The gut virome (the virus community of the gut) therefore has the potential to impact health and disease (e.g. cancer).[Comment: the gut virome isn't associted with the following diseases...] The human virome has already been associated with diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb]. The precedence of the virome impacting human health and the strong association of bacterial communities with colorectal cancer suggests that colorectal cancer may be associated with altered virus communities.

@@

Here we address the gap in knowledge of whether virus communities are altered in colorectal cancer and, if they are, how those changes might impact cancer progression and severity. We also aimed to evaluate their potential as a reservoir of diagnostic biomarkers. We report that colorectal cancer is associated with variation in colonic virus community composition. The majority of viruses identified within the virome were temperate bacteriophages. Just as the association between the bacterial community and colorectal cancer was driven by select influential bacteria including *Fusobacterium*, the association between the virome and colorectal cancer was driven by a subset of influential phages. Our data suggest that the influential phages do not exclusively infect influential bacterial, but rather act through the community as a whole. The implications of these findings are threefold. *First*, this supports a biological role for the virome in colorectal cancer development and suggests that more than bacteria are involved in the process. *Second*, we present a supplementary, or even alternative virus-based approach for classification modeling of colorectal cancer using stool samples. *Third*, we provide initial support for the importance of studying the virome as a component of the microbiome ecological network, especially in cancer. [Comment: I worry that from "We report..." to the end of the paragraph is too much a rehash of the Results. I would suggest moving this section to the abstract since the abstract is a bit light on results. ]

@@

Virus DNA was purified prior to sequencing, allowing us to analyze DNA exclusively from within virus capsids **(Figure \ref{sampleproc})**. Each extraction protocol was performed with a blank control to detect contaminants from reagents or other unintentional sources. Only one of the nine controls contained DNA at a minimal concentration (XX ng/ul), thus providing initial evidence of successful sequencing of VLP genomic DNA over potential contaminants **(Figure \ref{qualcontrol} A)**. As was expected, these controls yielded few sequences and were mostly removed while rarefying the datasets to a common number of sequences **(Figure \ref{qualcontrol} B)**. The high quality phage and bacterial sequences were assembled into highly covered contigs longer than 1kb **(Figure \ref{contigqc})**. Because contigs only represent genome fragments, we further clustered related contigs into operational genomic units (OGUs) **(Figure \ref{contigqc} - \ref{clustercontigqc})**.
[Comment: What do you think about using operational viral units (OVU) rather than OGU? I think of OGU as representing bacterial genomes. Adding phage to the name would make it more clear what we are refering to]

## Unaltered Virome Diversity in Colorectal Cancer
Microbiome and disease associations are often described as being of an altered diversity (i.e. "dysbiotic"). We therefore initially evaluated the changes in virome OGU diversity in colorectal cancer. We evaluated differences in communities between disease states using the Bray-Curtis, Shannon entropy, and richness diversity metrics. ~~To control for uneven sequencing depths, we rarefied to a minimum depth that maintained most samples while excluding the sparsely sequenced blank controls.~~ [Comment: put this in the methods section rarefied != sub-sampled, make sure you rarefied (i.e. subsampled 1000 times)] We observed no significant alterations in either Shannon entropy or richness in the diseased state **(Figure \ref{betaogu} C-D)**. There was also no statistically significant clustering of the disease groups (ANOSIM p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogu}**). It is worth noting that there was a significant difference between the few blank controls that remained after rarefying the data, and the other study groups (ANOSIM p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogunegative})**. This further supported the quality of our sample set. Similar to what was previously seen when analyzing the 16S rRNA gene sequences from the same samples [@Zackular:2014fba], standard alpha and beta diversity metrics were insufficient for capturing virus community differences between disease states.

@@

[Comment:
1A: Get rid of the legend title "Disease". I feel like the 0.1 and 0.9 marks on the ROC curves adds clutter without improving interpretation.
1B: Can you order: Bacteria, Virus, Combined, Metagenome - why is the combined AUC > virus when it seems the opposite in A? Can you put the AUCs in the legend for 1A and get rid of 1B?
1C: Format OTU names to be like "OTU 4" rather than "Otu00004"
1D: Format cluster  names to be like "Cluster 115" rather than "Cluster_115"
1E: See comments for 1C and 1D
Can you justify the y-axis in C, D, and E so they are vertically aligned and the x-axes are the same lengths?
]

@@

To determine the advantage of viral metagenomic methods over bacterial metagenomic methods, we compared the viral model to a model built using relative abundance profiles from bacterial metagenomic shotgun sequencing data. This model performed more poorly than any of the other models (mean AUC = `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = sigfig)`) **(Figure \ref{predmodel} A - B)**. Further investigation revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and low abundance OTUs **(Figure \ref{16scompare})**. Removal of OTUs with a median abundance of zero resulted in the removal of six OTUs, and a loss of model performance down to what was observed in the metagenome-based model **(Figure \ref{16scompare} A)**. The majority of these OTUs had a relative abundance lower than 1% **(Figure \ref{16scompare} B)**. [Comment: Can you add a comment about the nature of the features in the viral model?]

@@

After evaluating our ability to classify samples between two disease states, we performed a three-class random forest model including all disease states. The 16S rRNA gene model model yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Bacteria"),"meanauc"], digits = sigfig)` and outperformed the viral community model which yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Virus"),"meanauc"], digits = sigfig)` (p-value = `r signif(threesig, digits = sigfig)`, **Figure \ref{threewaymodel} A-C**). The microbes important for the cancer vs healthy and healthy vs adenoma models were also important for the three-class model **(Figure \ref{threewaymodel} D-E)**. The most important bacterium in the two and three class models was the same *Fusobacterium* (OTU 4) **(Figure \ref{predmodel} C, Figure \ref{threewaymodel} D)**. The viruses most important to the three-class model were also identified as bacteriophages **(Figure \ref{predmodel} D, Figure \ref{threewaymodel} E)**. {Influential phage relative abundances were both higher and lower in cancer **(Figure \ref{threewaymodel} F)**. Not all important OGUs were of increased abundance in the diseased state. The viral classification model depended on the unique signatures of these different abundance profiles to accurately classify each sample.} [Comment: I don't feel like the text between the {}'s says anything. It's like word salad.]

@@

## Diversity
Alpha and beta diversity were calculated using the operational genomic unit abundance profiles for each sample. Sequences were rarefied to 100,000 sequences. Samples with less than the cutoff were removed from the analysis. Alpha diversity was calculated using the Shannon Entropy and Richness metrics. Beta diversity was calculated using the Bray-Curtis metric (mean of 25 random sub-sampling iterations), and the statistical significance between the disease state clusters was assessed using an analysis of similarity (ANOSIM) with a post-hoc multivariate Tukey test. All diversity calculations were performed in R using the Vegan package [@veganCommunityEco:xdLliqSE.

@@

![*Results from healthy vs cancer classification models built using virome signatures, bacterial 16S rRNA gene sequence signatures, whole metagenomic signatures, and a combination of virome and 16S rRNA gene sequence signatures. A) ROC curve for visualizing the performance of each of the models for classifying stool as coming from either an individual with a cancerous or healthy colon. B) Quantification of the AUC variation for each model, and how it compared to each of the other models based on 15 iterations. A pairwise Wilcoxon test with a false discovery rate multiple hypothesis correction demonstrated that all models are significantly different from each other (p-value < 0.01). C) Mean decrease in accuracy (measurement of importance) of each operational taxonomic unit within the 16S rRNA gene classification model when removed from the classification model. Results based on 15 iterations. Mean is represented by a point, and bars represent standard error. D) Mean decrease in accuracy of each operational genomic unit in the virome classification model. E) Mean decrease in accuracy of each operational genomic unit and operational taxonomic unit in the model using both 16S rRNA gene and virome features.*\label{predmodel}](../figures/predmodel-viromebacteria.pdf){ width=100% }

>>>>

Colorectal cancer is the second leading cause of cancer-related death in the United States and is a primary cause of morbidity and mortality throughout the world. Although etiology for the majority of colorectal cancer cases remain unclear, its progression has been linked to changes in colonic bacterial community composition. Viruses are another important component of the colonic microbial community. However, they have yet to be studied in colorectal cancer, despite their oncogenic potential. We evaluated the colorectal cancer virome (virus community) using a cohort of 90 human subjects with either healthy, adenomatous (precancerous), or cancerous colons. We utilized 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing methods to compare the colorectal cancer virome to the bacterial community. We found that alpha and beta diversity metrics were insufficient for detecting virome changes in colorectal cancer, but more sophisticated random forest models identified striking changes in the virus community. The majority of the cancer-associated virome consisted of temperate bacteriophages, suggesting that the community was indirectly linked to colorectal cancer by modulating bacterial community structure and function. Our data suggested that the influential phages do not exclusively infect influential bacterial, but rather act through the community as a whole. These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and likely impact cancer progression by altering the bacterial host communities.

**Word Count**: 224 / 250

@@

# Introduction
Due to their mutagenic abilities and propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk], and because bacteria have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb], bacteriophages have the potential to indirectly impact cancer. The gut virome (the virus community of the gut) therefore has the potential to impact health and disease (e.g. cancer). Altered human virome composition and diversity have been identified in diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb]. The precedence of the virome impacting human health and the strong association of bacterial communities with colorectal cancer suggests that colorectal cancer may be associated with altered virus communities.

@@

Here we address the gap in knowledge of whether virus communities are altered in colorectal cancer and, if they are, how those changes might impact cancer progression and severity. We also aimed to evaluate their potential as a reservoir of diagnostic biomarkers. The implications of this study are threefold. *First*, this supports a biological role for the virome in colorectal cancer development and suggests that more than bacteria are involved in the process. *Second*, we present a supplementary, or even alternative virus-based approach for classification modeling of colorectal cancer using stool samples. *Third*, we provide initial support for the importance of studying the virome as a component of the microbiome ecological network, especially in cancer.

@@

Virus DNA was purified prior to sequencing, allowing us to analyze DNA exclusively from within virus capsids **(Figure \ref{sampleproc})**. Each extraction protocol was performed with a blank control to detect contaminants from reagents or other unintentional sources. Only one of the nine controls contained DNA, at a minimal concentration of 0.011 ng/µl, thus providing initial evidence of successful sequencing of VLP genomic DNA over potential contaminants **(Figure \ref{qualcontrol} A)**. As was expected, these controls yielded few sequences and were mostly removed while rarefying the datasets to a common number of sequences **(Figure \ref{qualcontrol} B)**. The high quality phage and bacterial sequences were assembled into highly covered contigs longer than 1kb **(Figure \ref{contigqc})**. Because contigs only represent genome fragments, we further clustered related contigs into operational genomic units (OGUs) **(Figure \ref{contigqc} - \ref{clustercontigqc})**.
[Comment: What do you think about using operational viral units (OVU) rather than OGU? I think of OGU as representing bacterial genomes. Adding phage to the name would make it more clear what we are refering to]

## Unaltered Virome Diversity in Colorectal Cancer
Microbiome and disease associations are often described as being of an altered diversity (i.e. "dysbiotic"). We therefore initially evaluated the changes in virome OGU diversity in colorectal cancer. We evaluated differences in communities between disease states using the Bray-Curtis, Shannon entropy, and richness diversity metrics. We observed no significant alterations in either Shannon entropy or richness in the diseased state **(Figure \ref{betaogu} C-D)**. There was also no statistically significant clustering of the disease groups (ANOSIM p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogu}**). It is worth noting that there was a significant difference between the few blank controls that remained after rarefying the data, and the other study groups (ANOSIM p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogunegative})**. This further supported the quality of our sample set. Similar to what was previously seen when analyzing the 16S rRNA gene sequences from the same samples [@Zackular:2014fba], standard alpha and beta diversity metrics were insufficient for capturing virus community differences between disease states.

@@

<!--
[Comment:
1A: Get rid of the legend title "Disease". I feel like the 0.1 and 0.9 marks on the ROC curves adds clutter without improving interpretation.
1B: Can you order: Bacteria, Virus, Combined, Metagenome - why is the combined AUC > virus when it seems the opposite in A? Can you put the AUCs in the legend for 1A and get rid of 1B?
1C: Format OTU names to be like "OTU 4" rather than "Otu00004"
1D: Format cluster  names to be like "Cluster 115" rather than "Cluster_115"
1E: See comments for 1C and 1D
Can you justify the y-axis in C, D, and E so they are vertically aligned and the x-axes are the same lengths?
]
-->

@@

To determine the advantage of viral metagenomic methods over bacterial metagenomic methods, we compared the viral model to a model built using relative abundance profiles from bacterial metagenomic shotgun sequencing data. This model performed more poorly than any of the other models (mean AUC = `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = sigfig)`) **(Figure \ref{predmodel} A - B)**. Further investigation revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and low abundance OTUs **(Figure \ref{16scompare})**. Removal of OTUs with a median abundance of zero resulted in the removal of six OTUs, and a loss of model performance down to what was observed in the metagenome-based model **(Figure \ref{16scompare} A)**. The majority of these OTUs had a relative abundance lower than 1% **(Figure \ref{16scompare} B)**. Although the features in the viral model were also of low abundance **(Figure \ref{threewaymodel} F)**, the coverage was sufficient for high model performance, likely because viral genomes are orders of magnitude smaller than bacterial genomes.

@@

After evaluating our ability to classify samples between two disease states, we performed a three-class random forest model including all disease states. The 16S rRNA gene model model yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Bacteria"),"meanauc"], digits = sigfig)` and outperformed the viral community model which yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Virus"),"meanauc"], digits = sigfig)` (p-value = `r signif(threesig, digits = sigfig)`, **Figure \ref{threewaymodel} A-C**). The microbes important for the cancer vs healthy and healthy vs adenoma models were also important for the three-class model **(Figure \ref{threewaymodel} D-E)**. The most important bacterium in the two and three class models was the same *Fusobacterium* (OTU 4) **(Figure \ref{predmodel} C, Figure \ref{threewaymodel} D)**. The viruses most important to the three-class model were also identified as bacteriophages **(Figure \ref{predmodel} D, Figure \ref{threewaymodel} E)**. Not all important OGUs were of increased abundance in the diseased state **(Figure \ref{threewaymodel} F)**.

@@

## Diversity
Alpha and beta diversity were calculated using the operational genomic unit abundance profiles for each sample. Sequences were rarefied to 100,000 sequences. Samples with less than the cutoff were removed from the analysis. Alpha diversity was calculated using the Shannon Entropy and Richness metrics. Beta diversity was calculated using the Bray-Curtis metric (mean of 25 random sub-sampling iterations), and the statistical significance between the disease state clusters was assessed using an analysis of similarity (ANOSIM) with a post-hoc multivariate Tukey test. All diversity calculations were performed in R using the Vegan package [@veganCommunityEco:xdLliqSE].

@@

![*Results from healthy vs cancer classification models built using virome signatures, bacterial 16S rRNA gene sequence signatures, whole metagenomic signatures, and a combination of virome and 16S rRNA gene sequence signatures. A) An example ROC curve for visualizing the performance of each of the models for classifying stool as coming from either an individual with a cancerous or healthy colon. B) Quantification of the AUC variation for each model, and how it compared to each of the other models based on 15 iterations. A pairwise Wilcoxon test with a false discovery rate multiple hypothesis correction demonstrated that all models are significantly different from each other (p-value < 0.01). C) Mean decrease in accuracy (measurement of importance) of each operational taxonomic unit within the 16S rRNA gene classification model when removed from the classification model. Results based on 15 iterations. Mean is represented by a point, and bars represent standard error. D) Mean decrease in accuracy of each operational genomic unit in the virome classification model. E) Mean decrease in accuracy of each operational genomic unit and operational taxonomic unit in the model using both 16S rRNA gene and virome features.*\label{predmodel}](../figures/predmodel-viromebacteria.pdf){ width=100% }
